---
output: github_document
---
pmetar
======

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

<!-- badges: start -->
[![CRAN](https://www.r-pkg.org/badges/version/pmetar)](https://cran.r-project.org/package=pmetar)
[![monthly](https://cranlogs.r-pkg.org/badges/pmetar)](https://www.rpackages.io/package/pmetar) 
[![total](https://cranlogs.r-pkg.org/badges/grand-total/pmetar)](https://www.rpackages.io/package/pmetar)
<!-- badges: end -->

Package allows to download current and historical METAR weather reports and decode basic parameters.

### METAR weather reports source web pages
The function metar_get scraps data from two web pages:

1. For current reports:

* Aviation Weather Center https://www.aviationweather.gov/metar</li><br>

2. For historical reports:

* Iowa Environmental Mesonet web page of Iowa State University
ASOS-AWOS-METAR http://mesonet.agron.iastate.edu/AWOS/</li>
* Weather Information Service http://www.ogimet.com/ developed by Guillermo Ballester Valor

Please take into consideration that the http://www.ogimet.com/ can block too frequent requests for data due to the server overload. <br>

### Locations of METAR stations
Information about the locations of the METAR stations / ariports were taken from two sources:

1. The first choice is the file from https://ourairports.com/data/ created by David Megginson.
2. If information can't be found, the second source is searched, the ADDS Station Table prepared by Greg Thompson NCAR/RAP. The file is available on the web page of Aviation Weather Center https://www.aviationweather.gov/docs/metar/stations.txt 
from NOAA National Weather Service https://www.weather.gov/

The both above data sources are in the public domain.

#### Package pmetar is for extracting and parsing information, from current or historical METAR reports, only.

#### Don't use it for flight planning or navigation!

### Installation

CRAN version

```{r eval=FALSE}
install.packages("pmetar")
```

Get the development version 0.2.7.25 from github:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("prcwiek/pmetar")
```

```{r}
library('pmetar')
```

## Contributors

* David Megginson, Author of data set with airports list https://ourairports.com/data/
* Greg Thompson, Author of data set with airports list https://www.aviationweather.gov/docs/metar/stations.txt

